The interactions of rational, pragmatic agents\\ lead to efficient language structure and use

### 2019-02-01 paper/paper.Rmd
update with NDGs last comments and BP walk through

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows: how does having property \textit{x} make using language $\ell$ more or less useful for communication? Among others, this efficiency-based framing has produced a rich set of theoretical and empirical targets exploring semantic typology [@RegierKempKay2015a], properties such as ambiguity [@Piantadosi2011a] and compositionality [@KirbyGriffithsSmith2014a], and the efficient use of reduction and redundancy in production [@LevyJaeger2007a; @GenzelCharniak2002a]. These projects typically employ information-theoretic principles to characterize efficiency in natural or simulated language corpora.\par

To preview, we diverge from this approach in two ways. First we derive a measure of efficiency from first principles of \textit{efficient language use} and show how this can lead to communicatively efficient systems. Second we demonstrate how assumptions about language-users impacts whether language properties are used efficiently. In this way, we bridge questions of design and use in a single framework.\par

@@

We provide a case study for this approach, in which functionalist regularities emerge from the dynamics of pragmatic communication. We choose a property of languages that could, in principle, vary freely, but shows strong regularities across languages. The explanadum is why this regularity holds. We examine ambiguity as our property, extending ideas by Piantadosi et al. [-@Piantadosi2011a]. We define a novel measure of efficiency that depends on the interactional behavior of speaker and listener agents. We adopt the reference game as our primary unit of interaction and model language users with the Rational Speech Act (RSA) framework -- a computional model of language use, which is nicely attested by experimental data on interaction. Using these basic ingredients, we show that the property of interest (ambiguity) is prevalent in languages that optimize our measure of efficiency. Further, we show \textit{how} ambiguity is used efficiently during local, in-the-moment interactions. The contributions of this work are twofold -- we derive a novel objective for measuring linguistic efficiency and also show how the reference game framework, in combination with formal models of communication, can be used to connect ideas about system-level efficiencies to in-the-moment language use.\par

@@

### Languages
A language $\ell$ defines the set of semantic mappings between utterance and meanings. For example, Figure 1 contains four utterances $U = \{\text{blue}, \text{green}, \text{square}, \text{circle}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}$. The boolean matrix describes the literal semantics of the language. We define a language as "ambiguous" if there is some utterance $u \in U$ which can apply to multiple meanings (i.e. $|[[u_i]]| > 1$)\footnote{We use double brackets $[[\dots]]$ to represent denotation.}. In Figure 1 both the words "square" and "green" are ambigous so we would say that $\ell$ contains ambiguity.

@@

An RSA *speaker agent* defines a conditional distribution over utterances, mapping from intended meanings $M$ to utterances $U$ using $\ell$ in a given context $c$. That is, a speaker defines $P_{speaker}(u|m, c; \ell)$. We will use $S(u|m, c; \ell)$ as short-hand throughout.  A *listener agent* defines a conditional distribution over meanings, mapping from utterances $U$ to meanings $M$ using $\ell$ in a given context $c$ (i.e. $L(m|u, c;\ell)$). Note that both speakers and listeners can induce joint distributions over utterace-meaning pairs, although, importantly, these distributions may differ:
$$P_{speaker}(u, m | c; \ell) = S(u|m, c; \ell)p(m|c)$$
$$P_{listener}(u, m| c; \ell) = L(m|u, c; \ell)p(u|c)$$

@@

Importantly, we assume that events $e = <u, m, c>$ are sampled according the to following generative model -- some context occurs in the world with probability $P(C=c)$. Within this context, an object $m$ occurs with probability $p(m|c)$. The speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m, c; \ell)$ (i.e. $e \sim p(c)p(m|c)S(u|m, c; \ell)$). From the ingredients described above, it is possible to derive the following objective between the speaker and listener distributions (see \href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 2.1} for a complete derivation).

\begin{equation}
\begin{split}
  = \mathbb{E}_{c \sim P(C)}[H_{cross}(P_{speaker}, P_{listener} | c; \ell)]\\
\end{split}
\end{equation}

From an information-theoretic perspective this objective is intuitive: $H_{cross}$ denotes the Cross-Entropy (CE), a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, we have an expectation over this term -- the expected difference between the distributions assumed by the speaker $P_{speaker}$ and listener $P_{listener}$ given a set of contexts $C$\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy.}. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.

@@

As an example, say we have two objects ($m_1$ and $m_2$), two utterances ($u_1$ and $u_2$), which differ in cost, and two languages ($\ell_1$ and $\ell_2$), which describe different utterance-meaning mappings. In language $\ell_1$, the low-cost $u_1$ can be used to refer to both $m_1$ and $m_2$ ($[[u_1]]_{\ell_1} = \{m_1, m_2 \}$), but the high-cost $u_2$ cannot be used at all ($[[u_2]]_{\ell_1} = \emptyset$). By contrast, in language $\ell_2$, $u_1$ can only refer to $m_1$ and $u_2$ can only refer to $m_2$ ($[[u_1]]_{\ell_2} = \{m_1\}$ and $[[u_2]]_{\ell_2} = \{m_2 \}$). While it is cheaper for a speaker to use $\ell_1$ (because speaking is always low-cost), it is more difficult for a listener (because $u_1$ is ambiguous). Crucially, however, if context is disambiguating then the speaker can use $u_1$ to refer to either $m_1$ or $m_2$ and $\ell_1$ should be preferred to $\ell_2$.\par

In the following experiments we explore two aspects of Piantadosi’s et al.'s claim. In Simulation 1, we examine the efficient language *structure* aspect of their claim, exploring when the optimal linguistic system $\ell^*$ is most likely to contain ambiguous expressions. In Simulation 2, we explore an efficient language *use* aspect of the claim -- at what point in a conversation is it useful for a speaker to use ambiguous expressions?\par

@@

While our results indicate that ambiguity is an efficient property of contextualized language, these simulations assumed that agents had perfect knowledge of the relevant conditional distributions ($P(M|C)$). This assumption may be too strong for describing much of day-to-day communication -- we seldom interact with others with perfect knowledge of the current context (or topic) at the start of a conversation. To explore how ambiguity may be \textit{used} efficiently in our framework, we next examine a case in which the listener has imperfect knowledge of context at the start of the conversation, but may infer it from the discourse history.\par

@@

In Simulation 1 we showed that efficiency defined in terms of pragmatic agents leads to a preference for languages that contain ambiguity. In Simulation 2 we assume a single fixed language $\ell$, which contains ambiguity, and instead vary the types of agents using $\ell$. We will show that efficient *use* of ambiguity depends on an agent's ability to use context for disambiguation. This instantiates an on-line perspective on the communicative function of ambiguity theory. More generally, it demonstrates how we can assess both questions of efficient use (current simulation) as well as design (previous simulation) in the same framework.\par

@@

Figure 3, panel (A) shows the empirical probability that a speaker uses an ambiguous utterance as a function of discourse position. The \textit{No pragmatic} baseline uses ambiguous utterances frequently and at a constant rate over the discourse. The \textit{Partial pragmatic} baseline avoids ambiguous utterances entirely. But, the \textit{Full pragmatic} model avoids ambiguous material only at the start of the discourse, employing it increasingly as the discourse proceeds. Panel (C) tracks our CE objective for each model over the discourse. Note that the objective decreases for all three models, primarily driven by the listener updating his belief about the actual topic ($P(C=c_{current}|D)$). However, the objective declines more quickly under the \textit{Full} and \textit{Partial} pragmatic speakers as listener agents are better able to infer the correct context. Additionally, the difference in CE between the \textit{Full}- and \textit{Partial pragmatic} models at the end of the discourse is driven by the reduction in speaker costs (see panel B). While speaker effort remains constant in both \textit{No pragmatic} and \textit{Partial pragmatic} baselines, effort declines in the \textit{Full pragmatic} model as she increasingly relies on ambiguous material later in the discourse.\par

@@

\vspace{1em} \fbox{\parbox[b][][c]{7.3cm}{\centering {Code and SI are available at this \href{https://github.com/benpeloquin7/zipf_principles/}{repository}.}}} \vspace{1em}
\noindent

>>>>

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows: how does having property \textit{x} make using language $\ell$ more or less useful for communication? Among others, this efficiency-based framing has produced a rich set of theoretical and empirical targets exploring semantic typology [@RegierKempKay2015a], properties such as ambiguity [@Piantadosi2011a] and compositionality [@KirbyGriffithsSmith2014a], and the efficient use of reduction and redundancy in production [@LevyJaeger2007a; @GenzelCharniak2002a].\par

The approaches above typically posit efficiency measures motivated by information-theoretic principles, but not directly grounded in language use by interacting agents. In this work, we derive a measure of efficiency from first principles of \textit{efficient language use} and show how this can lead to communicatively efficient systems. We also demonstrate how assumptions about interlocutors impacts how language properties are used efficiently. In this way, we bridge questions of design and use in a single framework.\par

@@

We provide a case study for this approach, in which functionalist regularities emerge from the dynamics of pragmatic communication. We choose a property of languages that could, in principle, vary freely, but shows strong regularities across languages. The explanadum is why this regularity holds. We examine ambiguity as our property, extending ideas by Piantadosi et al. [-@Piantadosi2011a]. We define a novel measure of efficiency that depends on the interactional behavior of speaker and listener agents. We adopt the reference game as our primary unit of interaction and model language users with the Rational Speech Act (RSA) framework -- a computional model of language use, which is nicely attested by experimental data on interaction. Using these ingredients, we show that the property of interest (ambiguity) is prevalent in languages that optimize our measure of efficiency (Simulation 1). Further, we show how ambiguity is used efficiently during local, in-the-moment interactions (Simulation 2). Put differently, these simulations examine efficiency from two angles -- in the first we vary languages, fixing agents, and search for efficient designs. In the second we vary agents, fixing language, and examine efficient use. \par

The contributions of this work are twofold -- we derive a novel measure of linguistic efficiency and also show how the reference game framework, in combination with formal models of communication, can be used to connect ideas about system-level efficiencies to in-the-moment language use.\par

@@

### Languages
A language $\ell$ defines the set of semantic mappings between utterances and meanings. For example, Figure 1 contains four utterances $U = \{\text{"blue"}, \text{"green"}, \text{"square"}, \text{"circle"}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}$. The boolean matrix describes the literal semantics of the language. We define a language as "ambiguous" if there is some utterance $u \in U$ which can apply to multiple meanings (i.e. $|[[u_i]]| > 1$)\footnote{We use double brackets $[[\dots]]$ to represent denotation.}. In Figure 1 both the words "square" and "green" are ambigous so we would say that $\ell$ contains ambiguity.

@@

An RSA *speaker agent* defines a conditional distribution over utterances, mapping from intended meanings $M$ to utterances $U$ using $\ell$ in a given context $c$. That is, a speaker defines $P_{speaker}(u|m, c; \ell)$. We will use $S(u|m, c; \ell)$ as short-hand throughout.  A *listener agent* defines a conditional distribution over meanings, mapping from utterances $U$ to meanings $M$ using $\ell$ in a given context $c$ (i.e. $L(m|u, c;\ell)$). Note that both speakers and listeners can induce joint distributions over utterance-meaning pairs, although, importantly, these distributions may differ:
$$P_{speaker}(u, m | c; \ell) = S(u|m, c; \ell)p(m|c)$$
$$P_{listener}(u, m| c; \ell) = L(m|u, c; \ell)p(u|c)$$

@@

Importantly, we assume that events $e = <u, m, c>$ are sampled according the to following generative model -- some context occurs in the world with probability $P(C=c)$. Within this context, an object $m$ occurs with probability $p(m|c)$. The speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m, c; \ell)$ (i.e. $e \sim p(c)p(m|c)S(u|m, c; \ell)$). From these ingredients it is possible to derive the following objective between the speaker and listener distributions (see \href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 2.1} for complete derivation).
\begin{equation}
\begin{split}
  = \mathbb{E}_{c \sim P(C)}[H_{cross}(P_{speaker}, P_{listener} | c; \ell)]\\
\end{split}
\end{equation}
From an information-theoretic perspective this objective is intuitive: $H_{cross}$ denotes the Cross-Entropy (CE), a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, we have an expectation over this term -- the expected difference between the distributions assumed by the speaker $P_{speaker}$ and listener $P_{listener}$ given a set of contexts $C$\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy.}. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.

@@

As an example, say we have two objects ($m_1$ and $m_2$), two utterances ($u_1$ and $u_2$), differing in cost, and two languages ($\ell_1$, $\ell_2$), describing different utterance-meaning mappings. In language $\ell_1$, the low-cost $u_1$ can be used to refer to both $m_1$ and $m_2$ ($[[u_1]]_{\ell_1} = \{m_1, m_2 \}$), but the high-cost $u_2$ cannot be used at all ($[[u_2]]_{\ell_1} = \emptyset$). By contrast, in language $\ell_2$, $u_1$ can only refer to $m_1$ and $u_2$ can only refer to $m_2$ ($[[u_1]]_{\ell_2} = \{m_1\}$ and $[[u_2]]_{\ell_2} = \{m_2 \}$). While it is cheaper for a speaker to use $\ell_1$ (because speaking is always low-cost), it is more difficult for a listener (because $u_1$ is ambiguous). Crucially, however, if context is disambiguating then the speaker can use $u_1$ to refer to either $m_1$ or $m_2$ and $\ell_1$ should be preferred to $\ell_2$.\par

In the following simulations we explore two aspects of Piantadosi’s et al.'s claim. In Simulation 1, we examine the efficient language *structure* aspect of their claim, exploring when the optimal linguistic system $\ell^*$ is most likely to contain ambiguous expressions. In Simulation 2, we explore an efficient language *use* aspect of the claim -- under what assumptions will agents use ambiguity efficiently in a conversation?\par

@@

While our results indicate that ambiguity is an efficient property of contextualized language use, these simulations assumed that agents had perfect knowledge of the relevant conditional distributions ($P(M|C)$). This assumption may be too strong for describing much of day-to-day communication -- we seldom interact with others with perfect knowledge of the current context (or topic) at the start of a conversation. To explore how ambiguity may be \textit{used} efficiently in our framework, we next examine a case in which the listener has imperfect knowledge of context at the start of the conversation, but may infer it from the discourse history.\par

@@

In Simulation 1 we showed that efficiency defined in terms of pragmatic agents leads to a preference for languages that contain ambiguity. In Simulation 2 we assume a single fixed language $\ell$, which contains ambiguity, and instead vary the types of agents using $\ell$. We will show that efficient *use* of ambiguity depends on an agent's ability to use context for disambiguation. More generally, Simulation 2 is intended to demonstrate how we can assess both questions of efficient use (current simulation) as well as design (previous simulation) in the same framework.\par

@@

Figure 3, panel (A) shows the empirical probability a speaker uses an ambiguous utterance as a function of discourse position. The \textit{No pragmatic} baseline uses ambiguous utterances frequently and at a constant rate over the discourse. The \textit{Partial pragmatic} baseline avoids ambiguous utterances entirely. But, the \textit{Full pragmatic} model avoids ambiguous material only at the start of the discourse, employing it increasingly as the discourse proceeds. Panel (C) tracks our CE objective for each model over the discourse. Note that the objective decreases for all three models, primarily driven by the listener updating his belief about the actual topic ($P(C=c_{current}|D)$). However, the objective declines more quickly under the \textit{Full} and \textit{Partial pragmatic} speakers as listener agents are better able to infer the correct context. Additionally, the difference in CE between the \textit{Full}- and \textit{Partial pragmatic} models at the end of the discourse is driven by the reduction in speaker costs. Panel (B) tracks speaker effort, which remains constant in both \textit{No pragmatic} and \textit{Partial pragmatic} baselines. But, effort declines in the \textit{Full pragmatic} model as she increasingly relies on ambiguous material later in the discourse.\par

@@

\vspace{1em} \fbox{\parbox[b][][c]{7.3cm}{\centering {Code and SI: \href{https://bit.ly/2DOnjI4}{\url{https://bit.ly/2DOnjI4}}.}}} \vspace{1em}
\noindent

### 2019-02-01 paper/paper.tex
update with NDGs last comments and BP walk through

Zipf articulated what is now considered a \emph{functionalist} approach
to language science -- analyzing language structure and use in terms of
efficiency. Such an approach might reframe our opening question as
follows: how does having property \textit{x} make using language
\(\ell\) more or less useful for communication? Among others, this
efficiency-based framing has produced a rich set of theoretical and
empirical targets exploring semantic typology (Regier, Kemp, \& Kay,
2015), properties such as ambiguity (Piantadosi, Tily, \& Gibson, 2011)
and compositionality (Kirby, Griffiths, \& Smith, 2014), and the
efficient use of reduction and redundancy in production (Genzel \&
Charniak, 2002; Levy \& Jaeger, 2007). These projects typically employ
information-theoretic principles to characterize efficiency in natural
or simulated language corpora.\par

To preview, we diverge from this approach in two ways. First we derive a
measure of efficiency from first principles of
\textit{efficient language use} and show how this can lead to
communicatively efficient systems. Second we demonstrate how assumptions
about language-users impacts whether language properties are used
efficiently. In this way, we bridge questions of design and use in a
single framework.\par

@@

We provide a case study for this approach, in which functionalist
regularities emerge from the dynamics of pragmatic communication. We
choose a property of languages that could, in principle, vary freely,
but shows strong regularities across languages. The explanadum is why
this regularity holds. We examine ambiguity as our property, extending
ideas by Piantadosi et al. (2011). We define a novel measure of
efficiency that depends on the interactional behavior of speaker and
listener agents. We adopt the reference game as our primary unit of
interaction and model language users with the Rational Speech Act (RSA)
framework -- a computional model of language use, which is nicely
attested by experimental data on interaction. Using these basic
ingredients, we show that the property of interest (ambiguity) is
prevalent in languages that optimize our measure of efficiency. Further,
we show \textit{how} ambiguity is used efficiently during local,
in-the-moment interactions. The contributions of this work are twofold
-- we derive a novel objective for measuring linguistic efficiency and
also show how the reference game framework, in combination with formal
models of communication, can be used to connect ideas about system-level
efficiencies to in-the-moment language use.\par

@@

A language \(\ell\) defines the set of semantic mappings between
utterance and meanings. For example, Figure 1 contains four utterances
\(U = \{\text{blue}, \text{green}, \text{square}, \text{circle}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}\).
The boolean matrix describes the literal semantics of the language. We
define a language as ``ambiguous'' if there is some utterance
\(u \in U\) which can apply to multiple meanings (i.e.
\(|[[u_i]]| > 1\))\footnote{We use double brackets $[[\dots]]$ to represent denotation.}.
In Figure 1 both the words ``square'' and ``green'' are ambigous so we
would say that \(\ell\) contains ambiguity.

@@

An RSA \emph{speaker agent} defines a conditional distribution over
utterances, mapping from intended meanings \(M\) to utterances \(U\)
using \(\ell\) in a given context \(c\). That is, a speaker defines
\(P_{speaker}(u|m, c; \ell)\). We will use \(S(u|m, c; \ell)\) as
short-hand throughout. A \emph{listener agent} defines a conditional
distribution over meanings, mapping from utterances \(U\) to meanings
\(M\) using \(\ell\) in a given context \(c\) (i.e. \(L(m|u, c;\ell)\)).
Note that both speakers and listeners can induce joint distributions
over utterace-meaning pairs, although, importantly, these distributions
may differ: \[P_{speaker}(u, m | c; \ell) = S(u|m, c; \ell)p(m|c)\]
\[P_{listener}(u, m| c; \ell) = L(m|u, c; \ell)p(u|c)\]

@@

Importantly, we assume that events \(e = <u, m, c>\) are sampled
according the to following generative model -- some context occurs in
the world with probability \(P(C=c)\). Within this context, an object
\(m\) occurs with probability \(p(m|c)\). The speaker attempts to refer
to that object by sampling from her conditional distribution
\(S(u|m, c; \ell)\) (i.e. \(e \sim p(c)p(m|c)S(u|m, c; \ell)\)). From
the ingredients described above, it is possible to derive the following
objective between the speaker and listener distributions (see
\href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 2.1}
for a complete derivation).

@@

As an example, say we have two objects (\(m_1\) and \(m_2\)), two
utterances (\(u_1\) and \(u_2\)), which differ in cost, and two
languages (\(\ell_1\) and \(\ell_2\)), which describe different
utterance-meaning mappings. In language \(\ell_1\), the low-cost \(u_1\)
can be used to refer to both \(m_1\) and \(m_2\)
(\([[u_1]]_{\ell_1} = \{m_1, m_2 \}\)), but the high-cost \(u_2\) cannot
be used at all (\([[u_2]]_{\ell_1} = \emptyset\)). By contrast, in
language \(\ell_2\), \(u_1\) can only refer to \(m_1\) and \(u_2\) can
only refer to \(m_2\) (\([[u_1]]_{\ell_2} = \{m_1\}\) and
\([[u_2]]_{\ell_2} = \{m_2 \}\)). While it is cheaper for a speaker to
use \(\ell_1\) (because speaking is always low-cost), it is more
difficult for a listener (because \(u_1\) is ambiguous). Crucially,
however, if context is disambiguating then the speaker can use \(u_1\)
to refer to either \(m_1\) or \(m_2\) and \(\ell_1\) should be preferred
to \(\ell_2\).\par

In the following experiments we explore two aspects of Piantadosi's et
al.'s claim. In Simulation 1, we examine the efficient language
\emph{structure} aspect of their claim, exploring when the optimal
linguistic system \(\ell^*\) is most likely to contain ambiguous
expressions. In Simulation 2, we explore an efficient language
\emph{use} aspect of the claim -- at what point in a conversation is it
useful for a speaker to use ambiguous expressions?\par

@@

While our results indicate that ambiguity is an efficient property of
contextualized language, these simulations assumed that agents had
perfect knowledge of the relevant conditional distributions
(\(P(M|C)\)). This assumption may be too strong for describing much of
day-to-day communication -- we seldom interact with others with perfect
knowledge of the current context (or topic) at the start of a
conversation. To explore how ambiguity may be \textit{used} efficiently
in our framework, we next examine a case in which the listener has
imperfect knowledge of context at the start of the conversation, but may
infer it from the discourse history.\par

@@

In Simulation 1 we showed that efficiency defined in terms of pragmatic
agents leads to a preference for languages that contain ambiguity. In
Simulation 2 we assume a single fixed language \(\ell\), which contains
ambiguity, and instead vary the types of agents using \(\ell\). We will
show that efficient \emph{use} of ambiguity depends on an agent's
ability to use context for disambiguation. This instantiates an on-line
perspective on the communicative function of ambiguity theory. More
generally, it demonstrates how we can assess both questions of efficient
use (current simulation) as well as design (previous simulation) in the
same framework.\par

@@

Figure 3, panel (A) shows the empirical probability that a speaker uses
an ambiguous utterance as a function of discourse position. The
\textit{No pragmatic} baseline uses ambiguous utterances frequently and
at a constant rate over the discourse. The \textit{Partial pragmatic}
baseline avoids ambiguous utterances entirely. But, the
\textit{Full pragmatic} model avoids ambiguous material only at the
start of the discourse, employing it increasingly as the discourse
proceeds. Panel (C) tracks our CE objective for each model over the
discourse. Note that the objective decreases for all three models,
primarily driven by the listener updating his belief about the actual
topic (\(P(C=c_{current}|D)\)). However, the objective declines more
quickly under the \textit{Full} and \textit{Partial} pragmatic speakers
as listener agents are better able to infer the correct context.
Additionally, the difference in CE between the \textit{Full}- and
\textit{Partial pragmatic} models at the end of the discourse is driven
by the reduction in speaker costs (see panel B). While speaker effort
remains constant in both \textit{No pragmatic} and
\textit{Partial pragmatic} baselines, effort declines in the
\textit{Full pragmatic} model as she increasingly relies on ambiguous
material later in the discourse.\par

@@

\fbox{\parbox[b][][c]{7.3cm}{\centering {Code and SI are available at this \href{https://github.com/benpeloquin7/zipf_principles/}{repository}.}}}
\vspace{1em} \noindent

>>>>

Zipf articulated what is now considered a \emph{functionalist} approach
to language science -- analyzing language structure and use in terms of
efficiency. Such an approach might reframe our opening question as
follows: how does having property \textit{x} make using language
\(\ell\) more or less useful for communication? Among others, this
efficiency-based framing has produced a rich set of theoretical and
empirical targets exploring semantic typology (Regier, Kemp, \& Kay,
2015), properties such as ambiguity (Piantadosi, Tily, \& Gibson, 2011)
and compositionality (Kirby, Griffiths, \& Smith, 2014), and the
efficient use of reduction and redundancy in production (Genzel \&
Charniak, 2002; Levy \& Jaeger, 2007).\par

The approaches above typically posit efficiency measures motivated by
information-theoretic principles, but not directly grounded in language
use by interacting agents. In this work, we derive a measure of
efficiency from first principles of \textit{efficient language use} and
show how this can lead to communicatively efficient systems. We also
demonstrate how assumptions about interlocutors impacts how language
properties are used efficiently. In this way, we bridge questions of
design and use in a single framework.\par

@@

We provide a case study for this approach, in which functionalist
regularities emerge from the dynamics of pragmatic communication. We
choose a property of languages that could, in principle, vary freely,
but shows strong regularities across languages. The explanadum is why
this regularity holds. We examine ambiguity as our property, extending
ideas by Piantadosi et al. (2011). We define a novel measure of
efficiency that depends on the interactional behavior of speaker and
listener agents. We adopt the reference game as our primary unit of
interaction and model language users with the Rational Speech Act (RSA)
framework -- a computional model of language use, which is nicely
attested by experimental data on interaction. Using these ingredients,
we show that the property of interest (ambiguity) is prevalent in
languages that optimize our measure of efficiency (Simulation 1).
Further, we show how ambiguity is used efficiently during local,
in-the-moment interactions (Simulation 2). Put differently, these
simulations examine efficiency from two angles -- in the first we vary
languages, fixing agents, and search for efficient designs. In the
second we vary agents, fixing language, and examine efficient use. \par

The contributions of this work are twofold -- we derive a novel measure
of linguistic efficiency and also show how the reference game framework,
in combination with formal models of communication, can be used to
connect ideas about system-level efficiencies to in-the-moment language
use.\par

@@

A language \(\ell\) defines the set of semantic mappings between
utterances and meanings. For example, Figure 1 contains four utterances
\(U = \{\text{"blue"}, \text{"green"}, \text{"square"}, \text{"circle"}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}\).
The boolean matrix describes the literal semantics of the language. We
define a language as ``ambiguous'' if there is some utterance
\(u \in U\) which can apply to multiple meanings (i.e.
\(|[[u_i]]| > 1\))\footnote{We use double brackets $[[\dots]]$ to represent denotation.}.
In Figure 1 both the words ``square'' and ``green'' are ambigous so we
would say that \(\ell\) contains ambiguity.

@@

An RSA \emph{speaker agent} defines a conditional distribution over
utterances, mapping from intended meanings \(M\) to utterances \(U\)
using \(\ell\) in a given context \(c\). That is, a speaker defines
\(P_{speaker}(u|m, c; \ell)\). We will use \(S(u|m, c; \ell)\) as
short-hand throughout. A \emph{listener agent} defines a conditional
distribution over meanings, mapping from utterances \(U\) to meanings
\(M\) using \(\ell\) in a given context \(c\) (i.e. \(L(m|u, c;\ell)\)).
Note that both speakers and listeners can induce joint distributions
over utterance-meaning pairs, although, importantly, these distributions
may differ: \[P_{speaker}(u, m | c; \ell) = S(u|m, c; \ell)p(m|c)\]
\[P_{listener}(u, m| c; \ell) = L(m|u, c; \ell)p(u|c)\]

@@

Importantly, we assume that events \(e = <u, m, c>\) are sampled
according the to following generative model -- some context occurs in
the world with probability \(P(C=c)\). Within this context, an object
\(m\) occurs with probability \(p(m|c)\). The speaker attempts to refer
to that object by sampling from her conditional distribution
\(S(u|m, c; \ell)\) (i.e. \(e \sim p(c)p(m|c)S(u|m, c; \ell)\)). From
these ingredients it is possible to derive the following objective
between the speaker and listener distributions (see
\href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 2.1}
for complete derivation).

@@

As an example, say we have two objects (\(m_1\) and \(m_2\)), two
utterances (\(u_1\) and \(u_2\)), differing in cost, and two languages
(\(\ell_1\), \(\ell_2\)), describing different utterance-meaning
mappings. In language \(\ell_1\), the low-cost \(u_1\) can be used to
refer to both \(m_1\) and \(m_2\)
(\([[u_1]]_{\ell_1} = \{m_1, m_2 \}\)), but the high-cost \(u_2\) cannot
be used at all (\([[u_2]]_{\ell_1} = \emptyset\)). By contrast, in
language \(\ell_2\), \(u_1\) can only refer to \(m_1\) and \(u_2\) can
only refer to \(m_2\) (\([[u_1]]_{\ell_2} = \{m_1\}\) and
\([[u_2]]_{\ell_2} = \{m_2 \}\)). While it is cheaper for a speaker to
use \(\ell_1\) (because speaking is always low-cost), it is more
difficult for a listener (because \(u_1\) is ambiguous). Crucially,
however, if context is disambiguating then the speaker can use \(u_1\)
to refer to either \(m_1\) or \(m_2\) and \(\ell_1\) should be preferred
to \(\ell_2\).\par

In the following simulations we explore two aspects of Piantadosi's et
al.'s claim. In Simulation 1, we examine the efficient language
\emph{structure} aspect of their claim, exploring when the optimal
linguistic system \(\ell^*\) is most likely to contain ambiguous
expressions. In Simulation 2, we explore an efficient language
\emph{use} aspect of the claim -- under what assumptions will agents use
ambiguity efficiently in a conversation?\par

@@

While our results indicate that ambiguity is an efficient property of
contextualized language use, these simulations assumed that agents had
perfect knowledge of the relevant conditional distributions
(\(P(M|C)\)). This assumption may be too strong for describing much of
day-to-day communication -- we seldom interact with others with perfect
knowledge of the current context (or topic) at the start of a
conversation. To explore how ambiguity may be \textit{used} efficiently
in our framework, we next examine a case in which the listener has
imperfect knowledge of context at the start of the conversation, but may
infer it from the discourse history.\par

@@

In Simulation 1 we showed that efficiency defined in terms of pragmatic
agents leads to a preference for languages that contain ambiguity. In
Simulation 2 we assume a single fixed language \(\ell\), which contains
ambiguity, and instead vary the types of agents using \(\ell\). We will
show that efficient \emph{use} of ambiguity depends on an agent's
ability to use context for disambiguation. More generally, Simulation 2
is intended to demonstrates how we can assess both questions of
efficient use (current simulation) as well as design (previous
simulation) in the same framework.\par

@@

Figure 3, panel (A) shows the empirical probability a speaker uses an
ambiguous utterance as a function of discourse position. The
\textit{No pragmatic} baseline uses ambiguous utterances frequently and
at a constant rate over the discourse. The \textit{Partial pragmatic}
baseline avoids ambiguous utterances entirely. But, the
\textit{Full pragmatic} model avoids ambiguous material only at the
start of the discourse, employing it increasingly as the discourse
proceeds. Panel (C) tracks our CE objective for each model over the
discourse. Note that the objective decreases for all three models,
primarily driven by the listener updating his belief about the actual
topic (\(P(C=c_{current}|D)\)). However, the objective declines more
quickly under the \textit{Full} and \textit{Partial} pragmatic speakers
as listener agents are better able to infer the correct context.
Additionally, the difference in CE between the \textit{Full}- and
\textit{Partial pragmatic} models at the end of the discourse is driven
by the reduction in speaker costs. Panel (B) tracks speaker effort,
which remains constant in both \textit{No pragmatic} and
\textit{Partial pragmatic} baselines. But, effort declines in the
\textit{Full pragmatic} model as she increasingly relies on ambiguous
material later in the discourse.\par

@@

\fbox{\parbox[b][][c]{7.3cm}{\centering {Code and SI: \href{https://bit.ly/2DOnjI4}{\url{https://bit.ly/2DOnjI4}}.}}}
\vspace{1em} \noindent
